\documentclass[10pt,letterpaper]{article}

\usepackage[T1]{fontenc}
\usepackage[margin=0.72in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{array}
\usepackage{booktabs}
\usepackage[dvipsnames]{xcolor}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{longtable}
\usepackage{tabularx}

\hypersetup{
  colorlinks=true,
  linkcolor=MidnightBlue,
  urlcolor=MidnightBlue,
  pdftitle={TileGrad: a typed tile language over tinygrad UOps},
}

\lstset{
  language=Python,
  basicstyle=\ttfamily\small,
  columns=fullflexible,
  keepspaces=true,
  frame=single,
  framerule=0.2pt,
  xleftmargin=0.4em,
  xrightmargin=0.4em,
  aboveskip=0.5em,
  belowskip=0.5em,
}

\setlist[itemize]{leftmargin=1.5em,itemsep=1pt,topsep=3pt}
\setlist[enumerate]{leftmargin=1.7em,itemsep=1pt,topsep=3pt}
\newcolumntype{Y}{>{\raggedright\arraybackslash}X}
\newcommand{\TileGrad}{\textsc{TileGrad}}
\newcommand{\tinygrad}{\textsc{tinygrad}}
\newcommand{\op}[1]{\textsc{#1}}
\newcommand{\must}{\textbf{MUST}}
\newcommand{\mustnot}{\textbf{MUST NOT}}
\newcommand{\should}{\textbf{SHOULD}}
\newcommand{\may}{\textbf{MAY}}

\title{\TileGrad: a typed tile language over \tinygrad{} UOps\\
\large Version 1 target specification}
\author{TileGrad contributors}
\date{July 2026}

\begin{document}
\maketitle

\begin{abstract}
\TileGrad{} is a statically scheduled, Python-embedded language for tiled
compute kernels.  Programs expose tile shapes, memory placement, launch shape,
and coarse scheduling intent while the compiler owns lane-level distribution,
synchronization, and legal instruction selection.  \TileGrad{} lowers into
\tinygrad{} UOps and delegates backend code generation, compilation, and
runtime execution to \tinygrad{}.

This document specifies the intended stable Version 1 language.  It is a
north-star contract rather than a description of the current experimental
implementation.  A conforming implementation may initially use scalar or SIMT
fallbacks for every tile operation.  Optimized copy, matrix, and pipeline
lowerings must preserve the same semantics.
\end{abstract}

\tableofcontents

\section{Status and Conformance}

The key words \must, \mustnot, \should, and \may{} describe normative
requirements.  Text marked ``non-normative'' provides rationale or examples.

A \emph{program} conforms when it uses only constructs defined by this
specification and satisfies all static validity rules.  An
\emph{implementation} conforms when it accepts every program required by its
declared profile, rejects invalid programs before device execution, and
preserves the observable semantics in this document.

Performance is not part of semantic conformance.  In particular, vectorized
loads, native matrix instructions, asynchronous transfers, physical pipeline
overlap, occupancy, and instruction counts are not guaranteed unless an
explicit capability is requested.

\subsection{Design goals}

\begin{itemize}
  \item Make tiled kernels substantially easier to express than raw custom
        \tinygrad{} UOps.
  \item Keep tile shape, memory placement, launch shape, and coarse scheduling
        visible in source and debug IR.
  \item Provide one correctness path whose meaning is independent of target
        instruction selection.
  \item Permit optimized lowering without changing source-level semantics.
  \item Keep the compiler boundary small: \TileGrad{} owns the typed tile IR;
        \tinygrad{} owns UOps, rendering, compilation, and runtime execution.
\end{itemize}

\subsection{Non-goals}

Version 1 does not standardize automatic schedule search, autotuning, arbitrary
layout functions, TileLang or TVM compatibility, backend source injection,
clusters, tensor memory, TMA, WGMMA, TCGEN05, user-managed asynchronous copy,
wait groups, warp specialization, scans, atomics, or implicit autograd.

\section{Compilation Model}

A source program passes through the following conceptual stages:

\begin{center}
\begin{tabularx}{\textwidth}{@{}lY@{}}
\toprule
\textbf{Stage} & \textbf{Responsibility} \\
\midrule
Specialize & Bind compile-time constants and tensor shape constants. \\
Parse & Convert the supported Python subset into source-located typed IR. \\
Verify & Check types, shapes, effects, access modes, uniformity, and resource rules. \\
Normalize views & Rewrite every view to a base buffer, offset, shape, and strides. \\
Legalize layouts & Select a canonical logical-parallel and fragment ownership layout. \\
Plan effects & Insert or validate dependencies and synchronization for managed memory. \\
Legalize tile ops & Select scalar/SIMT lowering or an exact available capability. \\
Lower & Emit scalar, vector, range, memory, barrier, and optional \op{Wmma} UOps. \\
Verify UOps & Apply the relevant \tinygrad{} UOp program specification. \\
Execute & Delegate rendering, compilation, launch, and storage to \tinygrad{}. \\
\bottomrule
\end{tabularx}
\end{center}

Each stage \must{} be inspectable.  At minimum, implementations expose the
parsed tile IR, verified tile IR, legalized tile IR, scalar/SIMT IR, and final
UOps.  A native operation selected during legalization must remain identifiable
in debug output.

\section{Python Frontend}

The canonical frontend is a Python decorator named
\texttt{@tilegrad.jit}.  It has one explicit behavior: calling a decorated
function specializes, compiles, caches, and executes a \TileGrad{} program.
An implementation \mustnot{} infer a different eager or lazy API from the
function body or return type.

\subsection{Supported source structure}

A decorated function consists of:

\begin{enumerate}
  \item runtime tensor and scalar parameters;
  \item keyword-only compile-time parameters annotated \texttt{T.Const};
  \item tensor shape declarations and output declarations;
  \item exactly one \texttt{with T.Kernel(...)} launch region;
  \item zero or more compile-time helper or macro expansions; and
  \item a return of every declared output, preserving tuple order.
\end{enumerate}

Each runtime tensor parameter must have exactly one annotated declaration of
the following form before its first use:
\begin{lstlisting}
name: T.Tensor[(shape...), dtype]
\end{lstlisting}
\texttt{T.const("M", ...)} declares named specialization symbols.  Every such
symbol must occur directly as one dimension in at least one input tensor
declaration.  At specialization, direct occurrences bind the symbol to the
corresponding runtime dimension; repeated occurrences must agree.  An
unbound symbol or conflicting dimensions are specialization errors.  Output
declarations do not bind symbols.

A runtime scalar parameter has an annotation \texttt{T.Scalar[dtype]}.  At the
current \tinygrad{} boundary it is represented by a dense rank-zero tensor
input and loaded once inside the kernel.  It cannot affect launch dimensions.
Compile-time helper functions execute during specialization.  Device macros
must be marked \texttt{@T.macro}, are non-recursive, and are inlined into the
typed IR before verification.

The device subset includes scalar assignments, tuple unpacking, calls to
specified \TileGrad{} operations, \texttt{for} over \TileGrad{} loop
constructs, structured \texttt{if}/\texttt{else}, a kernel launch
\texttt{with} statement, and a final return.  Python classes, recursion,
generators, exceptions, arbitrary iteration, \texttt{while}, \texttt{break},
and \texttt{continue} are outside Version 1.

Ordinary Python control flow may inspect only compile-time values.  A condition
depending on a device value creates device control flow.  A frontend
\mustnot{} coerce a device predicate to a Python boolean.

\subsection{Canonical program}

The following spelling is normative for the Version 1 decorator frontend.

\begin{lstlisting}
@tilegrad.jit
def matmul(A, B, *,
           BM: T.Const[int] = 64,
           BN: T.Const[int] = 64,
           BK: T.Const[int] = 32):
  M, N, K = T.const("M", "N", "K")
  A: T.Tensor[(M, K), T.float16]
  B: T.Tensor[(K, N), T.float16]
  C = T.empty((M, N), T.float16)

  with T.Kernel(T.ceildiv(N, BN), T.ceildiv(M, BM),
                threads=128) as (bx, by):
    A_shared = T.shared((BM, BK), T.float16)
    B_shared = T.shared((BK, BN), T.float16)
    accum = T.fragment((BM, BN), T.float32)
    T.clear(accum)

    for ko in T.Pipelined(T.ceildiv(K, BK), stages=3):
      T.copy(A.tile((by*BM, ko*BK), (BM, BK)),
             A_shared, pad=0)
      T.copy(B.tile((ko*BK, bx*BN), (BK, BN)),
             B_shared, pad=0)
      T.mma(A_shared, B_shared, accum)

    T.copy(accum, C.tile((by*BM, bx*BN), (BM, BN)))
  return C
\end{lstlisting}

\subsection{Normative API surface}

Version 1 implementations must accept the following source constructs.  Shape,
dtype, and integer arguments below are typed expressions subject to the static
rules in this document.

\begin{longtable}{@{}>{\raggedright\arraybackslash}p{6.3cm}>{\raggedright\arraybackslash}p{9.3cm}@{}}
\toprule
\textbf{Source form} & \textbf{Result or effect} \\
\midrule
\endhead
\texttt{T.const(*names)} & Declare specialization symbols; return one symbol
for one name and a tuple in argument order for multiple names. \\
\texttt{x: T.Tensor[(shape...), dtype]} & Declare the type of runtime tensor
parameter \texttt{x}. \\
\texttt{x: T.Scalar[dtype]} & Declare a runtime rank-zero scalar parameter. \\
\texttt{T.empty(shape, dtype)} & Declare and return a new global output tensor. \\
\texttt{T.shared(shape, dtype, sync="managed")} & Allocate workgroup shared
storage with managed or manual synchronization. \\
\texttt{T.fragment(shape, dtype, policy="auto")} & Allocate a collective
fragment using auto, row-major, or column-major policy. \\
\texttt{T.private(shape, dtype)} & Allocate private, statically indexed storage. \\
\texttt{V.tile(origin, shape)} & Create a possibly out-of-base tile view with
the same rank as \texttt{V}. \\
\texttt{V.subview(origin, shape)} & Create a view that must be provably within
the logical extent of \texttt{V}. \\
\texttt{V.transpose(order)} & Permute shape and strides by a rank permutation. \\
\texttt{V.reshape(shape)} & Create an equal-element-count view when contiguity
is proven. \\
\texttt{T.Kernel(*grid, threads=...)} & Enter the single launch region and
yield grid coordinates in dimension order. \\
\texttt{T.Serial(stop)} or \texttt{T.Serial(start, stop, step=1)} & Construct
an ordered half-open loop. \\
\texttt{T.Unroll(stop)} or \texttt{T.Unroll(start, stop, step=1)} & Construct
a compile-time-bounded loop requested to unroll. \\
\texttt{T.Reduce(stop)} or \texttt{T.Reduce(start, stop, step=1)} & Construct
an ordered scalar recurrence loop represented by a reduction axis. \\
\texttt{T.Parallel(*extents, policy="auto")} & Construct a logical Cartesian
parallel loop and yield one coordinate per extent. \\
\texttt{T.Pipelined(stop, stages=N)} or
\texttt{T.Pipelined(start, stop, stages=N)} & Construct a unit-step loop with
serial semantics and positive compile-time pipeline hint \texttt{N}. \\
\bottomrule
\end{longtable}

The scalar and tile operation forms are:

\begin{longtable}{@{}>{\raggedright\arraybackslash}p{6.3cm}>{\raggedright\arraybackslash}p{9.3cm}@{}}
\toprule
\textbf{Source form} & \textbf{Result or effect} \\
\midrule
\endhead
\texttt{T.load(V, indices, mask=True, other=0)} & Perform the scalar load
defined in Section~\ref{sec:scalar-memory}; \texttt{indices} is a scalar for
rank one and a tuple otherwise. \\
\texttt{T.store(V, indices, value, mask=True)} & Perform the scalar store
defined in Section~\ref{sec:scalar-memory}. \\
\texttt{V[indices]} and \texttt{V[indices] = value} & Abbreviate unmasked
\texttt{T.load} and \texttt{T.store}; fragment indexing is legal only in a
matching \texttt{T.Parallel}. \\
\texttt{T.fill(dst, value)} and \texttt{T.clear(dst)} & Fill a collective
destination with a scalar, or with zero. \\
\texttt{T.copy(src, dst, pad=0)} & Perform equal-shape synchronous copy. \\
\texttt{T.mma(A, B, C, transpose\_a=False, transpose\_b=False, mode="auto")}
& Accumulate a rank-two matrix product into \texttt{C}; mode is
\texttt{auto}, \texttt{simt}, or a versioned capability name. \\
\texttt{T.reduce(src, axis, op, accumulator\_dtype=None, keepdim=False,
initial=None, nan\_policy="propagate")} & Return a fresh fragment reduced by
\texttt{sum}, \texttt{max}, or \texttt{min}; omitted accumulator dtype means
the source dtype. \\
\texttt{T.reduce\_sum}, \texttt{T.reduce\_max}, \texttt{T.reduce\_min} &
Aliases fixing the corresponding \texttt{op} argument of \texttt{T.reduce}. \\
\texttt{T.where(predicate, x, y)} and \texttt{T.cast(x, dtype)} & Scalar or
pointwise lifted fragment select and conversion. \\
\texttt{T.thread\_id(dim)} and \texttt{T.thread\_extent(dim)} & Return expert
local-agent coordinates and extents. \\
\texttt{T.barrier()} & Perform the convergent workgroup barrier. \\
\bottomrule
\end{longtable}

Arithmetic, comparison, logical, and math operations automatically lift to
fragments and create the fragment-map nodes defined below.  Scalar-to-fragment
and compatible fragment-to-fragment broadcasting follows the Version 1
broadcast rule.  Rebinding a fragment name or assigning its logical coordinates
creates a new fragment version in typed IR.

\section{Parameters, Shapes, and Outputs}

\subsection{Value categories}

\begin{center}
\begin{tabularx}{\textwidth}{@{}lYY@{}}
\toprule
\textbf{Category} & \textbf{Meaning} & \textbf{Permitted uses} \\
\midrule
Compile-time constant & Included in the specialization cache key. & Python
control flow, tile and allocation shapes, workgroup shape, layouts, unrolling,
and all device expressions. \\
Tensor shape constant & Inferred from a runtime tensor, then specialized. &
Grid extents, output shapes, bounds, and device expressions. \\
Runtime scalar & Passed to the device program without specialization. & Device
expressions, predicates, indices, and loop bounds; not launch, allocation,
fragment, layout, unroll, or workgroup shapes. \\
\bottomrule
\end{tabularx}
\end{center}

Tensor rank and dtype are always specialization-time information.  Version 1
specializes tensor dimensions by default.  Implementations may later declare a
shape-polymorphic extension, but it must not weaken the rules for static shared
or fragment storage.

\subsection{Outputs}

\texttt{T.empty(shape, dtype)} declares a dense, contiguous global output.  It
does not allocate device-local storage inside the kernel.  Frontend lowering
creates an output specification and the host wrapper allocates the tensor
before invoking \tinygrad{} custom-kernel execution.

Every output \must{} be returned exactly once.  Only declared outputs may be
returned.  Output shape expressions may depend on input shape constants and
compile-time constants.  Multiple outputs are returned as a tuple in source
order.  Outputs must not alias inputs or one another.  Scalar device results
are represented as rank-zero or one-element output tensors, not Python values.

Runtime tensor inputs are read-only in the core profile.  A future in-place
extension must make read/write access explicit and must define aliasing.

\section{Scalar Types and Expressions}

Core scalar types are \texttt{bool}, signed and unsigned 8-, 16-, 32-, and
64-bit integers, \texttt{float16}, \texttt{bfloat16}, and
\texttt{float32}.  An implementation may expose additional \tinygrad{} dtypes
through declared capabilities.

Expressions are statically typed.  Implicit conversion is limited to a Python
literal converted to the unambiguous type required by its context.  All other
dtype changes require \texttt{cast}.  Integer indices use the implementation's
index type and must not be silently converted from floating point.

\begin{center}
\begin{tabularx}{\textwidth}{@{}lY@{}}
\toprule
\textbf{Class} & \textbf{Operations} \\
\midrule
Arithmetic & \texttt{+}, \texttt{-}, \texttt{*}, floating division, integer
floor division, remainder, negation, minimum, and maximum. \\
Comparison & \texttt{<}, \texttt{<=}, \texttt{==}, \texttt{!=},
\texttt{>=}, and \texttt{>}; result dtype is \texttt{bool}. \\
Logical & elementwise and, or, not, and \texttt{select(predicate, a, b)}. \\
Conversion & numeric cast and bitcast where equal storage width is proven. \\
Math & absolute value, reciprocal, square root, \texttt{exp2},
\texttt{log2}, and sine. \\
Constants & finite numeric values, positive and negative infinity, and typed
reduction identities. \\
\bottomrule
\end{tabularx}
\end{center}

\texttt{exp} and \texttt{log} may be frontend compositions over
\texttt{exp2} and \texttt{log2}.  Floating-point reassociation and reduction
order are implementation-defined within the numerical contract in
Section~\ref{sec:numerics}.

\section{Buffers and Views}

A buffer type is the tuple
\[
  (\mathrm{dtype},\;\mathrm{rank},\;\mathrm{shape},\;\mathrm{strides},
    \;\mathrm{address\ space},\;\mathrm{access}).
\]

Core implementations support ranks zero through four.  They may support higher
ranks.  External tensors are dense, non-overlapping, and contiguous.  This
restriction matches the current \tinygrad{} custom-kernel boundary, which may
materialize non-contiguous input views.

A view is
\[
  V = (B,\;o,\;\mathbf{s},\;\mathbf{d},\;n_B),
\]
where \(B\) is a base buffer, \(o\) is an element offset,
\(\mathbf{s}\) is a logical shape, and \(\mathbf{d}\) contains element
strides.  The base storage contains \(n_B\) elements.  The address of logical
coordinate \(\mathbf{i}\) is
\[
  \mathrm{addr}(V,\mathbf{i}) = o + \sum_k i_k d_k.
\]

A coordinate is in bounds exactly when
\[
  \mathrm{inbounds}(V,\mathbf{i}) \iff
  \left(\bigwedge_k 0\leq i_k<s_k\right)
  \land 0\leq\mathrm{addr}(V,\mathbf{i})<n_B.
\]
The storage extent is the element count of an external tensor or allocation.
\texttt{tile(origin, shape)} may intentionally create a logical view extending
beyond base storage; masked load, store, fill, and copy use the predicate above
to define edge behavior.

Subview and transpose are view operations and do not copy.  Reshape is valid
only when contiguity can be proven.  Writable zero-stride views, overlapping
writable views, and negative strides are invalid in Version 1.

Two views overlap when their sets of valid base-storage addresses intersect.
Copy between identical views is a no-op.  Every other overlapping source and
destination copy is invalid; Version 1 does not provide snapshot or
\texttt{memmove} semantics.  If disjointness cannot be proven, compilation
must fail.

Negative coordinates are out of bounds; they do not wrap as Python indices.
Bounds behavior is defined by the operation performing the access.  There is
no general pass that silently repairs arbitrary out-of-bounds shared or private
accesses.

\section{Memory Spaces}

\begin{center}
\begin{tabularx}{\textwidth}{@{}lYYY@{}}
\toprule
\textbf{Space} & \textbf{Ownership} & \textbf{Shape} & \textbf{Purpose} \\
\midrule
Global & Invocation/tensor & Static rank; specialized dimensions & Inputs and outputs. \\
Shared & Workgroup & Compile-time constant & Cooperative staging and communication. \\
Fragment & Workgroup collective & Compile-time constant & Logically shaped values distributed across lanes. \\
Private & One local agent & Compile-time constant & Scalars and statically indexed register tuples. \\
\bottomrule
\end{tabularx}
\end{center}

Shared, fragment, and private allocations have lexical lifetime inside one
launch region.  Dynamic shared allocation is outside the core profile.
Private tuple indexing must become constant after specialization and unrolling;
otherwise compilation fails rather than silently spilling to another memory
space.

\section{Execution Model}

A launch is a grid of independent workgroups.  A grid has one to three positive
specialization-time extents.  A workgroup has one to three positive
specialization-time local extents or one positive total thread count.  No order
or implicit synchronization exists between workgroups.

The values yielded by \texttt{T.Kernel} are grid coordinates.  Expert code
uses \texttt{T.thread\_id(dim)} for a local coordinate and
\texttt{T.thread\_extent(dim)} for its extent, where \texttt{dim} is
\(0\), \(1\), or \(2\).

\subsection{Loop forms}

\begin{center}
\begin{tabularx}{\textwidth}{@{}lY@{}}
\toprule
\textbf{Form} & \textbf{Semantics} \\
\midrule
Serial & Ordered iterations over a half-open integer interval. \\
Unroll & Serial semantics with a compile-time request to expand iterations. \\
Reduce & Ordered recurrence semantics that the lowerer may reassociate when
allowed by the numerical contract. \\
Parallel & An unordered logical Cartesian iteration space; every point executes
exactly once. \\
Pipelined & Serial logical semantics with a pipeline-depth optimization hint. \\
\bottomrule
\end{tabularx}
\end{center}

\subsection{Logical parallel mapping}

\texttt{T.Parallel} is not an alias for a hardware local axis.  It defines an
unordered logical iteration space distributed over the current workgroup.  The
compiler selects from a bounded set of canonical policies.  Let the iteration
shape be \((e_0,\ldots,e_{r-1})\), let \(P\) be the workgroup size, and let
\(\ell\in[0,P)\) be the row-major linear local-agent identifier.

\begin{itemize}
  \item \texttt{row\_major}: use
        \(q_R(\mathbf{i})=\sum_k i_k\prod_{j>k}e_j\);
        owner \(q_R\bmod P\), private slot \(\lfloor q_R/P\rfloor\);
  \item \texttt{column\_major}: use
        \(q_C(\mathbf{i})=\sum_k i_k\prod_{j<k}e_j\);
        owner \(q_C\bmod P\), private slot \(\lfloor q_C/P\rfloor\); and
  \item an internal exact mapping required by a selected native matrix capability.
\end{itemize}

An agent executes the coordinates whose linearized value is
\(\ell+tP\) for nonnegative \(t\) while that value is less than
\(\prod_k e_k\).  This rule covers iteration spaces smaller than, equal to, or
larger than the workgroup.

The source may request a canonical policy or \texttt{auto}.  \texttt{auto}
may select only from the specified finite set.  Unsupported or ambiguous
composition must fail deterministically.  The selected policy \must{} appear
in legalized debug IR.

If explicit thread identifiers affect the same memory operations as a logical
parallel loop, the loop policy must be explicit.  An implementation must not
make observable program behavior depend on an unreported automatic mapping.

\section{Fragments}

A fragment is a statically shaped, workgroup-collective logical value.  Its
physical register ownership is not observable source state.  Legalization
assigns each logical coordinate to a canonical lane and private slot, or to an
exact native matrix fragment layout.

Every fragment carries a resolved mapping policy.  A logical parallel region
``matches'' a fragment when its extents equal the fragment shape and its
resolved policy is identical.  Access to coordinate \(\mathbf{i}\) in that
region is executed only by the owner defined by the mapping formula above.
All operations produce a new logical fragment version; source versions are read
before destination writes, including syntactically in-place elementwise forms.

Fragments support:

\begin{itemize}
  \item fill and clear;
  \item synchronous copy to and from compatible views;
  \item elementwise unary, binary, comparison, select, and cast operations;
  \item rank-compatible scalar and row/column broadcasting;
  \item axis reductions; and
  \item matrix multiply-accumulate.
\end{itemize}

Fragment coordinates may be read or written only by a tile operation or inside
a matching logical parallel region.  Arbitrary fragment indexing by a runtime
thread identifier is invalid.  SIMT fallback must implement the same logical
coordinate ownership; it must not replicate the complete fragment in every
thread.

\subsection{Fragment elementwise algebra}

\op{MapUnary}\((f,X)\), \op{MapBinary}\((f,X,Y)\),
\op{Select}\((P,X,Y)\), and \op{Cast}\((T,X)\) apply pointwise and produce a
fresh fragment.  Operands use right-aligned broadcasting: for each aligned
axis, extents must be equal or one; a missing leading axis and an extent-one
axis repeat the same source coordinate.  The result extent is the maximum of
the aligned extents.  Dtypes follow the scalar expression rules.

Scalar operands are rank-zero fragments for broadcasting purposes.  Row and
column broadcasting are not special cases: shapes \((M,1)\) and \((1,N)\)
follow the same rule.  Fragment-to-fragment copy requires equal shapes; use an
elementwise broadcast operation when expansion is intended.

\section{Loads, Stores, and Effects}\label{sec:scalar-memory}

The scalar memory operations are:

\begin{align*}
  \op{Load}(V,\mathbf{i},m,a) &=
    \begin{cases}V[\mathbf{i}] & m\land\mathrm{inbounds}(V,\mathbf{i})\\
                  a & \text{otherwise,}\end{cases}\\
  \op{Store}(V,\mathbf{i},x,m) &: \quad
    V[\mathbf{i}]\leftarrow x\quad\text{iff }m\land\mathrm{inbounds}(V,\mathbf{i}).
\end{align*}

The mask defaults to true.  The alternate load value defaults to typed zero.
A value is explicitly cast before storage when source and destination dtypes
differ.

Effects to the same logical location are ordered by source order unless they
occur in an unordered parallel region.  Two unordered writes, or an unordered
read and write without a defined collective or synchronization relation, form
a data race and are invalid.  Version 1 has no atomic escape from this rule.

\section{Tile Operations}

Tile operations are collective statements with explicit logical read and write
regions.  They must be reached uniformly by every participating local agent.
A tile operation in thread-divergent control flow is invalid unless the
operation explicitly defines predicated collective participation.

\subsection{Fill and clear}

\[
  \op{Fill}(D,x):\qquad D[\mathbf{i}]\leftarrow
  \mathrm{cast}_{D.\mathrm{dtype}}(x)
\]
for every valid logical coordinate \(\mathbf{i}\) in \(D\).
\op{Clear}\((D)\) is exactly \op{Fill}\((D,0)\).  Fill never writes outside
the destination view.

\subsection{Synchronous copy}

For equal-shaped logical regions \(S\) and \(D\),
\[
  \op{Copy}(S,D,p):\qquad
  D[\mathbf{i}]\leftarrow
  \begin{cases}
    \mathrm{cast}_{D.\mathrm{dtype}}(S[\mathbf{i}]) &
      \mathrm{inbounds}(S,\mathbf{i})\\
    \mathrm{cast}_{D.\mathrm{dtype}}(p) & \text{otherwise}
  \end{cases}
\]
when \(D[\mathbf{i}]\) is in bounds.  An out-of-bounds destination coordinate
performs no write.

Source and destination shapes must be equal.  Copy does not perform implicit
broadcasting and does not infer a tile extent from an unrelated operand.

Copy has safe-to-consume semantics: a subsequent dependent high-level tile
operation observes the completed destination values.  The compiler inserts or
elides transfer completion and workgroup visibility operations as necessary.
This guarantee includes ordering before managed shared storage is reused.

The compiler may lower copy with scalar, vector, SIMT-cooperative, or
target-specific transfer instructions, but instruction choice is not
observable.  A regular copy never exposes a partially completed asynchronous
transfer.

\subsection{Matrix multiply-accumulate}

For rank-two operands, \op{Mma} has the mathematical semantics
\[
  C \leftarrow C + \mathrm{cast}_{T_{acc}}(\mathrm{op}_A(A))
                    \times
                    \mathrm{cast}_{T_{acc}}(\mathrm{op}_B(B)),
\]
where each \(\mathrm{op}\) is identity or transpose.  Shapes must satisfy
\[
  A:(M,K),\qquad B:(K,N),\qquad C:(M,N)
\]
after transpose.  The operation does not implicitly clear \(C\).

Core implementations support float32 SIMT MMA and float16 or bfloat16 inputs
with float32 accumulation.  Additional dtype pairs are capabilities.

The selection mode is one of:

\begin{itemize}
  \item \texttt{auto}: use any conforming SIMT or native implementation;
  \item \texttt{simt}: do not select a native matrix instruction; or
  \item an exact native capability: require its atom shape, dtypes, participant
        count, and layout, failing compilation if unavailable.
\end{itemize}

An exact request must never silently fall back.  \texttt{auto} must preserve
the same logical fragment and numerical contract whichever path is selected.

An exact native capability declares an atom \((M_a,N_a,K_a)\).  A logical
MMA may use that capability only when \(M\), \(N\), and \(K\) are integer
multiples of the corresponding atom dimensions.  The compiler partitions
\(M\) and \(N\) into disjoint atoms and visits \(K\) atoms in increasing
logical order.  A global edge is represented by already padded input tiles and
suppressed output stores; an exact native MMA itself never accesses a partial
atom.  If these rules cannot be met, an exact request fails and \texttt{auto}
uses another legal capability or SIMT.

\subsection{Axis reduction}

\op{Reduce}\((X,a,f,T_{acc},k)\) reduces axis \(a\) of \(X\), using
operation \(f\), accumulator dtype \(T_{acc}\), and keep-dimension flag \(k\).
Version 1 operations are sum, maximum, and minimum.

The result shape removes axis \(a\) when \(k\) is false and replaces it with
one when \(k\) is true.  Reduction starts from the operation identity unless
an explicit initial value is provided.  Values are converted to the accumulator
dtype before combination.

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Operation} & \textbf{Identity} & \textbf{NaN policy} \\
\midrule
sum & typed zero & NaNs propagate \\
maximum & negative infinity or integer minimum & explicit propagate or ignore \\
minimum & positive infinity or integer maximum & explicit propagate or ignore \\
\bottomrule
\end{tabular}
\end{center}

Maximum and minimum default to \texttt{nan\_policy="propagate"}.  An
implementation may emulate this policy when a native instruction differs.
Reduction order is otherwise implementation-defined.

\section{Synchronization and Uniformity}

Managed high-level operations use effect analysis to establish required
completion and visibility.  The implementation must conservatively preserve:

\begin{itemize}
  \item read-after-write dependencies;
  \item write-after-read dependencies before storage reuse;
  \item write-after-write source order; and
  \item loop-carried dependencies.
\end{itemize}

A portable workgroup barrier waits for all local agents and establishes a
happens-before relation from prior shared writes to subsequent shared reads by
participants.  Every barrier must be reached convergently by the same
participant set.  A barrier under a thread-varying conditional is invalid.

Expert code may request
\texttt{T.shared(shape, dtype, sync="manual")} instead of the default
\texttt{sync="managed"}.  Managed
and manual synchronization must not be mixed for one allocation.  Raw shared
access through explicit thread identifiers is permitted only for manually
synchronized allocations; the author is then responsible for barriers and
race freedom.  \texttt{T.barrier()} is the source operation for the portable
workgroup barrier.  High-level collective copy, fill, reduction, and MMA do not
accept manually synchronized shared operands in the core profile.

There is no grid-wide barrier in the core profile.

\section{Pipelined Loops}

A pipelined loop has exactly the observable behavior of dependency-respecting
serial iterations.  Each logical iteration executes once, each effect occurs
once, and no consumer observes uninitialized or prematurely overwritten
storage.

The stage count is a positive compile-time optimization hint.  It does not
guarantee physical overlap, a number of in-flight iterations, a particular
instruction, a particular number of storage versions, or improved performance.
Executing the loop serially is semantically conforming.

An optimized implementation may reorder operations, version shared storage,
and use asynchronous target instructions internally.  It must preserve the
synchronous copy contract and all loop-carried dependencies.  Manual
stage/order arrays, transfer tokens, wait groups, and hardware barrier parity
are not Version 1 source constructs.

\section{Structured Control Flow}

Structured \texttt{if}/\texttt{else} is permitted over a boolean scalar.
Expression-level selection should be used for simple predication.  Loads and
stores may carry explicit masks.

A conditional is \emph{workgroup-uniform} when every local agent evaluates the
same branch.  Collective tile operations and barriers may appear only in
uniform control flow.  The compiler must prove uniformity from compile-time
values, grid coordinates, and other known-uniform values, or reject the
program.  Local thread identifiers and values loaded under thread-varying
indices are conservatively non-uniform.  Version 1 has no user assertion that
overrides uniformity analysis.

\section{Numerical Contract}\label{sec:numerics}

Integer addition, subtraction, and multiplication are modulo \(2^w\) for a
\(w\)-bit dtype.  Comparisons use the dtype's signedness.  Floor division and
remainder follow mathematical floor division.  Defined programs have a
nonzero divisor and shift counts in \([0,w)\).  Statically invalid cases must be
rejected; behavior outside the defined domain is not covered by conformance.

Floating-point elementwise operations follow the selected \tinygrad{} dtype
and backend within ordinary IEEE-754 expectations.  The compiler may fuse a
multiply and add, reassociate a reduction, or use a native matrix instruction.
Consequently, semantic equality is not generally bitwise equality.

Conformance vectors use finite inputs in \([-1,1]\) unless exceptional values
are the subject of the test.  Integer, boolean, movement, mask, and correctly
rounded cast tests are exact.  Default floating tolerances are:

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Class} & \textbf{Absolute tolerance} & \textbf{Relative tolerance} \\
\midrule
float32 elementwise & \(10^{-6}\) & \(10^{-6}\) \\
float32 reduction/MMA, length \(R\) & \(10^{-5}\max(1,R)\) & \(10^{-5}\) \\
float16/bfloat16 output & \(2\cdot10^{-2}\) & \(2\cdot10^{-2}\) \\
FlashAttention output & \(2\cdot10^{-2}\) & \(2\cdot10^{-2}\) \\
\bottomrule
\end{tabular}
\end{center}

A native and SIMT MMA must implement the same mathematical accumulation dtype,
but may differ in rounding within these limits.  NaNs compare classification
only because payload and sign may be canonicalized.  Infinity and signed-zero
tests compare classification and sign exactly.  Selected reduction NaN-policy
tests compare whether a NaN is present.  Masked-off values must not introduce
exceptional values beyond the specified alternate or reduction identity.

\section{Capabilities and Failures}

Capabilities describe optional legal lowerings, not alternate language
semantics.  A capability record includes a stable name, target predicate,
operand dtypes, accumulator dtype, atom shape, participant count, operand
scope requirements, and internal fragment layout.

Capability names are namespaced and versioned, for example
\texttt{tinygrad.cuda.m16n8k16.f16.f32.v1}.  The implementation exposes
\texttt{tilegrad.capabilities(target)} to enumerate records.  Each record also
contains the exact \tinygrad{} \op{Wmma} argument schema expected by the
compatibility adapter.  Renaming or changing the meaning of a record requires
a new versioned name.

The compiler must distinguish:

\begin{itemize}
  \item an invalid \TileGrad{} program;
  \item a valid program unavailable on the selected profile or target;
  \item a failure to produce valid \tinygrad{} UOps; and
  \item a backend compilation or runtime failure.
\end{itemize}

Diagnostics must identify the source construct, expected contract, observed
types or shapes, and selected target where relevant.  Unsupported exact native
requests fail during capability legalization, before UOp construction.

\section{Boundary with \tinygrad{}}

\TileGrad{} owns the frontend, typed IR, specialization, views, effects,
uniformity, canonical logical layouts, synchronization planning, tile
legalization, SIMT algorithms, and exact native capability adaptation.

\tinygrad{} owns UOp validation and rewriting after legalization, scalar and
vector rendering, device-specific code generation, compilation, launch,
storage integration, and outer tensor-graph scheduling.

\TileGrad{} must not depend on \tinygrad{} optimization to recover missing
memory dependencies, fragment ownership, or synchronization.  Integration
occurs through one versioned compatibility layer and a tested \tinygrad{}
revision range.

The core profile has no implicit gradient.  Differentiation through a
\TileGrad{} kernel must fail clearly unless an explicit backward kernel or VJP
is registered.

\section{Conformance Profiles}

\subsection{Core profile}

The core profile requires:

\begin{itemize}
  \item decorator parsing and specialization;
  \item dense contiguous tensor inputs and one or more outputs;
  \item one- through three-dimensional grids and workgroups;
  \item scalar expressions, views, masks, and structured conditionals;
  \item managed shared storage and canonical logical parallel mapping;
  \item fragments, fill, clear, copy, sum/max/min reduction, and SIMT MMA;
  \item serial execution of a semantically pipelined loop when necessary; and
  \item complete debug IR and final UOp validation.
\end{itemize}

\subsection{Accelerated profile}

The accelerated profile is additive.  It may provide vectorized copy, native
matrix capabilities, real software pipelining, asynchronous instructions hidden
behind synchronous operations, and additional dtypes.  A backend may implement
some accelerated capabilities without implementing all of them.

\section{Required Conformance Programs}

A release claiming core conformance must test at least:

\begin{enumerate}
  \item exact and masked one- through four-dimensional copy;
  \item internal strided subviews and transpose;
  \item source padding, destination suppression, cross-dtype copy, and rejected
        overlapping copies;
  \item one- through three-dimensional grids, canonical ownership mappings,
        logical parallel loops, and rejected runtime-scalar launch extents;
  \item uniform and rejected divergent collective operations;
  \item shared-memory producer/consumer and reuse dependencies;
  \item fragment unary, binary, broadcast, select, and cast operations;
  \item sum, maximum, and minimum reductions, including non-power-of-two tails;
  \item float32 SIMT GEMM with M, N, and K edge tiles;
  \item float16 and bfloat16 input GEMM with float32 accumulation;
  \item unavailable exact-native capability failures;
  \item multiple output allocation and return ordering; and
  \item small causal and non-causal FlashAttention forward kernels using stable
        online softmax, edge sequence lengths, reduction, \texttt{exp2}, and
        two MMA operations.
\end{enumerate}

The same mathematical cases should run through SIMT and every available native
capability.  Negative tests are part of conformance, not optional robustness
tests.

\appendix

\section{Normative IR Summary}

The exact in-memory Python classes are not normative.  A conforming typed IR
must represent at least the following semantic nodes.

\begin{longtable}{@{}>{\raggedright\arraybackslash}p{3.0cm}>{\raggedright\arraybackslash}p{4.6cm}>{\raggedright\arraybackslash}p{8.0cm}@{}}
\toprule
\textbf{Node} & \textbf{Operands} & \textbf{Meaning} \\
\midrule
\endhead
Program & parameters, outputs, launch & One specialized callable program. \\
Buffer parameter & type, shape, access & Dense external tensor storage. \\
Scalar parameter & dtype, category & Compile-time or runtime scalar. \\
Output specification & shape, dtype, order & Host-allocated returned tensor. \\
View & base, offset, shape, strides, storage extent & Logical region sharing base storage. \\
Allocate shared & shape, dtype, sync mode & Workgroup-visible storage. \\
Allocate fragment & shape, dtype & Collective logical register tile. \\
Allocate private & shape, dtype & Per-agent statically indexed storage. \\
Kernel launch & grid, workgroup & Grid of independent workgroups. \\
Serial loop & start, stop, step, body & Ordered iteration. \\
Unroll loop & static interval, body & Ordered iteration requested to expand. \\
Reduce loop & interval, recurrence & Scalar recurrence. \\
Parallel loop & extents, policy, body & Unordered logical iteration. \\
Pipelined loop & interval, stages, body & Serial semantics with pipeline hint. \\
Thread index & dimension & Explicit local-agent coordinate or extent. \\
If & predicate, branches & Structured control flow. \\
Load & view, coordinate, mask, alternate & Predicated scalar read. \\
Store & view, coordinate, value, mask & Predicated scalar write. \\
Fill & destination, scalar & Collective typed fill. \\
Copy & source, destination, pad & Collective synchronous region copy. \\
Fragment map & operation, sources, broadcast shape & Pointwise collective fragment operation. \\
MMA & A, B, C, transpose, mode & Collective accumulating matrix multiply. \\
Reduce & source, axis, op, dtype, keepdim & Collective axis reduction. \\
Barrier & memory scope & Convergent workgroup synchronization. \\
\bottomrule
\end{longtable}

\section{Non-normative Implementation Milestones}

The intended implementation order is:

\begin{enumerate}
  \item typed parameters, outputs, scalar expressions, and normalized views;
  \item the decorator parser and source-located diagnostics;
  \item canonical logical parallel mapping and fragment ownership;
  \item first-class fill and copy with conservative synchronization planning;
  \item fragment elementwise algebra and axis reductions;
  \item mixed-precision SIMT MMA and the FlashAttention conformance kernel;
  \item one exact native MMA capability adapted to \tinygrad{} \op{Wmma}; and
  \item optimized pipeline rewriting after shared-memory effects are stable.
\end{enumerate}

The current builder frontend may remain as an internal construction and testing
tool, but the Version 1 public contract is the decorated typed language defined
above.

\end{document}
